---
title: "G2F_xgb_predictions"
author: "Jim Holland"
date: "2023-01-13"
output: html_document
---

Based on xgboost grid search, select some parameters, refit the xgb on the full training data set and predict 2022 values. Check correlations among predictions from different models.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
#library(xgboost)
```

# Get metadata and submission template
```{r}
template = read.csv("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Testing_Data/1_Submission_Template_2022.csv")

g2f = fread(file = "Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Jim/G2F_EC477_geno500_ge1k.csv")
metaCols = names(g2f)[1:8]
#make boolean vectors of test observations
valid = g2f$Year == 2022
valid.meta = setDF(copy(g2f))[valid,c(metaCols)] %>%
  select(Env, Hybrid)
```

# Get predictions from first model
```{r}
preds1 = read.table("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Weighted_predictions/xgb_wt_pred2_10_0.2_0.1_0.25")
```

```{r}
hist(apply(preds1, 2, var))
```
What are the lowest variance values
```{r}
sort(apply(preds1, 2, var))[1:10]
```
Probably reasonable

# Get predictions from 2nd model
```{r}
preds2 = read.table("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Weighted_predictions/xgb_wt_pred2_10_0.3_0.5_0.75")
```

```{r}
hist(apply(preds2, 2, var))
```

What are the lowest variance values
```{r}
sort(apply(preds2, 2, var))[1:10]
```

# Get predictions from 3rd model
```{r}
preds3 = read.table("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Weighted_predictions/xgb_wt_pred2_10_0.5_0.5_0.5")
```

```{r}
hist(apply(preds3, 2, var))
```

What are the lowest variance values
```{r}
sort(apply(preds3, 2, var))[1:10]
```

# Get predictions from 4th model
```{r}
preds4 = read.table("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Weighted_predictions/xgb_wt_pred2_100_0.2_0.25_0.25")
```

```{r}
hist(apply(preds4, 2, var))
```

What are the lowest variance values
```{r}
sort(apply(preds4, 2, var))[1:10]
```

# Make the ensemble as average of all

Maybe we should just make an average of all as an ensemble:
```{r}
ensemble = apply(do.call(cbind, list(preds1, preds2, preds3, preds4)), 1, mean)
hist(ensemble)
```

# Create a submission file for these predictions:
We need to put our predictions is the correct order.
```{r}
valid.meta$Yield_Mg_ha = ensemble
template$Yield_Mg_ha = NULL
submit1 = left_join(template, valid.meta) %>% 
  select(Env, Hybrid, Yield_Mg_ha)

#check that order hasn't changed
all(template$Env == submit1$Env)
```

Another check on order
```{r}
all(template$Hybrid == submit1$Hybrid)
```
OK, order is good.

Check how these predictions compare to the unweighted predictions
```{r}
unwt = read.csv("Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Ensemble_predictions/xgb_4model_ensemble_predictions.csv")
plot(unwt$Yield_Mg_ha, submit1$Yield_Mg_ha)
```
They're quite similar
```{r}
cor(unwt$Yield_Mg_ha, submit1$Yield_Mg_ha)
```
Check the mean predicted yields by environment
```{r}
submit1 %>% group_by(Env) %>%
  summarize(mean.yield = mean(Yield_Mg_ha))
```

```{r}
write.csv(submit1, "Q:/.shortcut-targets-by-id/14BgjSlCN-5Nt_TAwg9-2nM6ZZrbkSh4y/G2F_Final_Competition_Data/Weighted_predictions/xgb_wt_4model_ensemble_predictions.csv", 
          row.names = F, quote = F)
```
