---
title: "Analysis of G2F Trait Data"
author: "Qiuyue Chen"
date: "December 1, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'X:/g2fcompetition2022') #set to the current working directory
```

```{r}
library(tidyverse)
```

Read in data, remove all local checks and any hybrids missing from vcf.
```{r}
df <- read.csv("./Training_Data/1_Training_Trait_Data_2014_2021.csv")
#Totally 136012 observations

Hybrids <- read.csv("./Training_Data/All_hybrid_names_info.csv")

df2 <- df[df$Hybrid %in% Hybrids[Hybrids$vcf == "TRUE", 'Hybrid'], ]
#Currently 131433 observations
colnames(df2)
```

Convert some columns to factor for plotting.
```{r}
df2 = df2 %>%
  mutate(
    Env = as.factor(Env),
    #Field_Location = as.factor(Field_Location),
    Stand_per_area = Stand_Count_plants / Plot_Area_ha,
    id = 1:n() #make an index for observations for filtering later
  )
```

Check Stand_Count_plants within Env.
```{r}
year.mns = group_by(df2, Env) %>%
  summarise(Stand_Count_plants = mean(Stand_Count_plants, na.rm = T))

no.data = year.mns %>% filter(is.na(Stand_Count_plants))  
no.data
```

set the count for these Env mannually based on previous years or personal communications.
```{r}
df2[df2$Env == "MNH1_2016", 'Stand_Count_plants'] <- 70
df2[df2$Env == "MNH1_2018", 'Stand_Count_plants'] <- 70
df2[df2$Env == "NEH1_2015", 'Stand_Count_plants'] <- 70
df2[df2$Env == "NEH4_2015", 'Stand_Count_plants'] <- 70
df2[df2$Env == "TXH2_2018", 'Stand_Count_plants'] <- 70
df2[df2$Env == "TXH4_2019", 'Stand_Count_plants'] <- 70
df2[df2$Env == "WIH3_2020", 'Stand_Count_plants'] <- 70
```


Check distribution of yield within Env.
```{r}
year.mns = group_by(df2, Env) %>%
  summarise(Yield_Mg_ha = mean(Yield_Mg_ha, na.rm = T)) 

no.data = year.mns %>% filter(is.na(Yield_Mg_ha))  
no.data
```

Plot the relationship between Yield and stand over all years.
```{r}
yield.stand = ggplot(df2, aes(x = Stand_per_area, y = Yield_Mg_ha)) +
  geom_point(size = 0.3, alpha = 0.2, colour = 'green') +
  ggtitle("Relationship between Yield and Stand_per_area") +
  theme_bw()
yield.stand
#ggsave(yield.stand, filename = "yield.stand.png", width = 15, height = 15, units = "cm")
```

Now filter. First filter by Stand_Count_plants.
```{r}
#identify outliers as 2.5*IQR and/or as above/below 30% and plot them
stand.plot = ggplot(df2, aes(x = Env, y = Stand_Count_plants)) +
  geom_boxplot(outlier.size = 1) +
  ggtitle('Distribution of Stand_Count_plants within each Env') + 
  stat_summary(fun = mean, geom = "point", shape = 23, size = 1) +
  theme(legend.position = "bottom", legend.key.size = unit(0.15, 'cm'), legend.text = element_text(size = 5), 
  axis.text.x= element_text(angle = 60, hjust = 1, colour = "black", size = 5)) 

outliers = group_by(df2, Env) %>% 
  summarise(Mean.std = mean(Stand_Count_plants, na.rm=T), 
            IQR.std = IQR(Stand_Count_plants, na.rm=T)) %>%
  mutate(top.iqr = Mean.std + 2.5*IQR.std,
         bot.iqr = Mean.std - 2.5*IQR.std, 
         top.perc = Mean.std*1.3,
         bot.perc = Mean.std*0.7)

outliers.iqr = merge(outliers, df2, by = "Env", all = T) %>%
  filter(Stand_Count_plants < bot.iqr | Stand_Count_plants > top.iqr)

outliers.perc = merge(outliers, df2, by = "Env", all = T) %>%
  filter(Stand_Count_plants < bot.perc | Stand_Count_plants > top.perc)

outliers.both = rbind(outliers.iqr, outliers.perc) %>% 
  filter(duplicated(.)) #keep only rows that are duplicated in both!

#stand.plot + geom_point(data = outliers.iqr, colour = "red", size = 0.5) +
#    ggtitle("Outliers based on 2.5 IQR")

#stand.plot + geom_point(data = outliers.perc, colour = "green", size = 0.5) +
#    ggtitle("Outliers based on percentage of mean")

stand.plot + geom_point(data = outliers.both, colour = "magenta", size = 0.5) +
    ggtitle("Outliers based on percentage of mean AND IQR")

#Remove all of the rows where the stand count is an outlier. 
df2 <- df2[!(df2$id %in% outliers.both$id), ]
#remove 1704 observations -> 129729 observations remain
```

Filter for Yield_Mg_ha.
```{r}
yield.plot = ggplot(df2, aes(x = Env, y = Yield_Mg_ha)) +
  geom_boxplot(outlier.size = 1) +
  ggtitle('Distribution of Yield_Mg_ha within each Env') + 
  stat_summary(fun = mean, geom = "point", shape = 23, size = 1) +
  theme(legend.position = "bottom", legend.key.size = unit(0.15, 'cm'), legend.text = element_text(size = 5), 
  axis.text.x= element_text(angle = 60, hjust = 1, colour = "black", size = 5)) 

outliers = group_by(df2, Env) %>% 
  summarise(Mean.std = mean(Yield_Mg_ha, na.rm=T), 
            IQR.std = IQR(Yield_Mg_ha, na.rm=T)) %>%
  mutate(top.iqr = Mean.std + 2.5*IQR.std,
         bot.iqr = Mean.std - 2.5*IQR.std, 
         top.perc = Mean.std*1.3,
         bot.perc = Mean.std*0.7)

outliers.iqr = merge(outliers, df2, by = "Env", all = T) %>%
  filter(Yield_Mg_ha < bot.iqr | Yield_Mg_ha > top.iqr)

outliers.perc = merge(outliers, df2, by = "Env", all = T) %>%
  filter(Yield_Mg_ha < bot.perc | Yield_Mg_ha > top.perc)

outliers.both = rbind(outliers.iqr, outliers.perc) %>% 
  filter(duplicated(.)) #keep only rows that are duplicated in both!

#yield.plot + geom_point(data = outliers.iqr, colour = "red", size = 0.5) +
#    ggtitle("Outliers based on 2.5 IQR")

#yield.plot + geom_point(data = outliers.perc, colour = "green", size = 0.5) +
#    ggtitle("Outliers based on percentage of mean")

yield.plot + geom_point(data = outliers.both, colour = "magenta", size = 0.5) +
    ggtitle("Outliers based on percentage of mean AND IQR")  
	
#Remove all of the rows where yield is an outlier.
df2 <- df2[!(df2$id %in% outliers.both$id),]
#remove 741 observations -> 128988 observations remain
```

```{r}
#Remove all of the rows where yield is missing. remove 5783 observations -> 123205 observations remain
df2 <- df2[!is.na(df2$Yield_Mg_ha), ]
df2 <- df2[, colnames(df2) != c('id', 'Stand_per_area')]

write.csv(df2, file = "./1_Training_Trait_Data_2014_2021_filtered.csv", row.names = FALSE, quote = FALSE)
```

